---
title: "R Notebook"
author: "Orlando Sabogal"
output: 
  html_notebook: 
    highlight: zenburn
    number_sections: yes
    theme: yeti
    toc: yes
---

```{r}
library(tidyverse)
library(sf)
library(leaflet)
```

Leaflet is a javascript [library](https://leafletjs.com/). If yoy know your way around **js** it goes something like this: 

```{r,eval=FALSE}
##This is javascript code, DO NOT RUN!

var map = L.map('map').setView([51.505, -0.09], 13);

L.tileLayer('https://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png', {
    attribution: '&copy; <a href="https://www.openstreetmap.org/copyright">OpenStreetMap</a> contributors'
}).addTo(map);

L.marker([51.5, -0.09]).addTo(map)
    .bindPopup('A pretty CSS3 popup.<br> Easily customizable.')
    .openPopup();
```

You should also know to configure a CCS file and include the Leaflet library in the header section of a html file:

```{r, eval=FALSE}
##This is html code, DO NOT RUN!

<link rel="stylesheet" href="https://unpkg.com/leaflet@1.5.1/dist/leaflet.css"
   integrity="sha512-xwE/Az9zrjBIphAcBb3F6JVqxf46+CDLwfLMHloNu6KEQCAWi6HcDUbeOfBIptF7tcCzusKFjFw2yuvEpDL9wQ=="
   crossorigin=""/>

     
 ##<!-- Make sure you put this AFTER Leaflet's CSS -->
 
     <script src="https://unpkg.com/leaflet@1.5.1/dist/leaflet.js"
   integrity="sha512-GffPMF3RvMeYyc1LWMHtK8EbPv0iNZ8/oTtHPx9/cc2ILxQ+u905qIwdpULaqDkyBKgOaB57QTMg7ztg8Jm2Og=="
   crossorigin=""> </script>

```

Fortunately, in the [html widgets for R](http://www.htmlwidgets.org/) there is a [Leaflet for R](https://rstudio.github.io/leaflet/) that let you use most of the capabilities of Leaflet with your regular R code. To create maps with leaflet you can use *data.frame* objects, *sf* objects, *raster* objects and *json* files. 

Leaflet for R is methodologically easy to use and you can get farther by exploring the arguments of the functions. A nice feature of Leaflet is that is integrated with shiny. 

<br/>
<hr/>

#Rudiments

To create a leaflet object you only have to call the **leaflet()** function. If you have an object (think about an *sf*) which you want to plot you should pass it through the function as **leaflet( -YourObject- )**. To add layers the **%>%** operator is used before a new function defining a new characteristic of the map. 
As a background layer leaflet enables to use different maps providers with **addProviderTiles** or **addTiles.**

```{r}
leaflet() %>% 
  addProviderTiles(providers$Esri.NatGeoWorldMap)
```

You can see the Tiles options [here](http://leaflet-extras.github.io/leaflet-providers/preview/). 
In the previous map you can zoom in and zoom out all around the world. You can set the specific view you want with the **setView()** function as in the following examples where the maps are centred in *Plaza de la indepencia* (Montevideo, Uruguay)

```{r}
leaflet() %>% 
  addProviderTiles(providers$Esri.NatGeoWorldMap) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 8)

```

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16)
```



Of course, the main benefit of using leaflet is getting your own data into the map. You can add a marker with **addMarkers** and specifying the *long* and *lat* arguments 

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addMarkers(-56.199735, -34.906543)
```

You can even assign a pop-up by setting the *popup* argument.

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addMarkers(-56.199735, -34.906543, popup = "Plaza Independencia")
```

Add many more markers can be a matter of calling the **addMarker** function multiple times. 

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addMarkers(-56.199735, -34.906543, popup = "Plaza Independencia") %>% 
  addMarkers(-56.198309, -34.906563, popup = "Palacio Salvo")
```


Off course, it does not make sense to add geographic information one by one; rather, a better option is give to the function an object and automatically plot all the markers.

```{r}
Sites <- data.frame(longitud = c(-56.199735, -56.198309, -56.201037), 
                    lattitud = c(-34.906543, -34.906563, -34.907896), 
                    Name = c("Plaza Independencia", "Palacio Salvo", "Teatro Solis"))

leaflet(Sites) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addMarkers(lng = ~longitud, lat = ~lattitud, popup = ~Name)

```

In the last example we configured our data in the **leaflet()** function, but is also possible to do it in the **addMarkers()** function. What do you think is the difference?

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addMarkers(data = Sites, lng = ~longitud, lat = ~lattitud, popup = ~Name)
```


A marker can also be a circle:

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addCircleMarkers(data = Sites, lng = ~longitud, lat = ~lattitud)
```

<br/>
<hr/>

#Points and Circles

But is time to plot more natural "geometric-vectorial" elements as a circle (**addCircles()**)

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addCircles(data = Sites, lng = ~longitud, lat = ~lattitud,
                   radius = 100)
```


What is the difference among **addCircleMarker()** and **addCircle().**?

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addCircles(data = Sites, lng = ~longitud, lat = ~lattitud,
                   radius = 100, opacity = 1, fillOpacity = 1)
```

Within the arguments of the function there are option to set elements like the radius, the opacity or the colour. 

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -56.199735, lat = -34.906543, zoom = 16) %>% 
  addCircles(data = Sites, lng = ~longitud, lat = ~lattitud,
                   radius = 100, opacity = 1, fillOpacity = 0.7, color = "black", fillColor = "green")
```

<br/>
<hr/>

#Using our shapes

```{r}
setwd("~/Dropbox/Teaching/SpatialAnalysis-MontevideoWorkshop2019/Notebooks/Montevideo_Data/Vectoriales_2011")
Uruguay <- st_read("ine_depto.shp")
```

```{r}
setwd("~/Dropbox/Teaching/SpatialAnalysis-MontevideoWorkshop2019/Notebooks/Montevideo_Data/Accidentes2006-2010")
Accidents <- st_read("accidentes2006-2010.shp")
```

```{r}
Uruguay = st_set_crs(Uruguay, 32721)
Accidents = st_set_crs(Accidents, 32721)

Uruguay <- st_transform(Uruguay,4326)
Accidents = st_transform(Accidents, 4326)
```

Intuitively, perhaps you already know that to get our Uruguay map we only need to use the **leaflet()** function, pass the **sf** object and then use the specific function to create areas. This function is **addPolygons().**

```{r}
leaflet(Uruguay) %>% 
  addPolygons()
```

And then again, a tile can provide some context to your map. 

```{r}
leaflet(Uruguay) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons()
```

```{r}
leaflet(Uruguay) %>% 
  addProviderTiles(providers$Stamen.Watercolor) %>% 
  addPolygons()
```


```{r}
leaflet(Uruguay) %>% 
  addProviderTiles(providers$Stamen.Watercolor) %>% 
  addPolygons(color = "black", weight = 2, fillColor = "white")
```

A taste of the power of Leaflet is in highlighting the polygons of the shape and perform some operations. 

```{r}
leaflet(Uruguay) %>% 
  addProviderTiles(providers$Stamen.Watercolor) %>% 
  addPolygons(color = "black", weight = 2, fillColor = "white", fillOpacity = 1,
              highlightOptions = highlightOptions(color = "blue", fillColor = "blue", 
                                                  weight = 0, bringToFront = TRUE))
```

```{r}
leaflet(Uruguay) %>% 
  addProviderTiles(providers$Stamen.Watercolor) %>% 
  addPolygons(color = "black", weight = 2, fillColor = "white", fillOpacity = 1,
              label = ~NOMBRE, labelOptions = labelOptions(
                style = list("font-weight" = "normal"), 
                textsize = "10px"),
              highlightOptions = highlightOptions(color = "blue", fillColor = "blue", 
                                                  weight = 0, bringToFront = TRUE))
```


After you get used to leaflet you realize that is easy to create awesome maps. You may struggle a little  with the inside function-arguments and probably will iterate a few times, but you will get it done. 

```{r}
#source: https://rstudio.github.io/leaflet/choropleths.html &
# From http://leafletjs.com/examples/choropleth/us-states.js

#install.packages("geojsonio")

#states <- geojsonio::geojson_read("json/us-states.geojson", what = "sp")
#see: https://github.com/rstudio/leaflet/issues/498

states <- geojsonio::geojson_read( 
        x = "https://raw.githubusercontent.com/PublicaMundi/MappingAPI/master/data/geojson/us-states.json"
        , what = "sp"
    )


bins <- c(0, 10, 20, 50, 100, 200, 500, 1000, Inf)
pal <- colorBin("YlOrRd", domain = states$density, bins = bins)

labels <- sprintf(
  "<strong>%s</strong><br/>%g people / mi<sup>2</sup>",
  states$name, states$density
) %>% lapply(htmltools::HTML)

leaflet(states) %>%
  setView(-96, 37.8, 4) %>%
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.light",
    accessToken = Sys.getenv('MAPBOX_ACCESS_TOKEN'))) %>%
  addPolygons(
    fillColor = ~pal(density),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(pal = pal, values = ~density, opacity = 0.7, title = NULL,
    position = "bottomright")


```


```{r}
Accidents2009 <- Accidents %>% filter(ANIO == "2009")
leaflet(Accidents2009) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircles()
```



