---
title: "Warm Up"
output: 
  html_notebook: 
    number_sections: yes
    toc: yes
---

<br/> 
In this document I will show some basic functions from the **sf** package so you can have the intuition of how it works and what kind of operations can be performed. You will learn how to *read* geographic data into your R workflow and how to *write* geographic data in you computer. In addition you will see a workflow analysis and the map making process. <br/>
You cand download all files used from my [Github Repository](https://github.com/orlando-sabogal/SpatialAnalysis-MontevideoWorkshop2019). Please notice that all datasets are in the folder */Notebooks/Montevideo_Data*.

```{r}
library(sf)
library(tmap)
library(tidyverse)
```

```{r}
setwd("~/Dropbox/Teaching/SpatialAnalysis-MontevideoWorkshop2019/Notebooks/Montevideo_Data")
#The path to the working directory should change according to where you allocated the repository in your computer. 

#setwd('..')
```

```{r}
getwd()
```


#Points (Accidents)

The most common type for spatial datasets are ESRI shape files. You can use the *st_read()* function and set the *dsn* argument (data source name).

```{r}

setwd("~/Dropbox/Teaching/SpatialAnalysis-MontevideoWorkshop2019/Notebooks/Montevideo_Data/Accidentes2006-2010")
Accidents <- st_read("accidentes2006-2010.shp")

```

The above function created the object *Accidents*. As usual, you can print the object in your console (or rmarkdown document) by typing the name of the object or get a general overview with **str()** and **summary()** . An easy way to make a map is by passing the *sf* object to the **plot()** function.

```{r}
Accidents
```

**Question:** Why are not the 41121?

```{r}
str(Accidents)
```


```{r}
summary(Accidents)
```

```{r}
plot(Accidents)
```

By default, **plot()** creates map for every variable within the file. Use **st_geometry** to get only the points. 

```{r}
plot(st_geometry(Accidents))
```

```{r}
plot(st_geometry(Accidents), cex = 0.05, col = "blue")
```

You can add some settings to the maps. Nevetheless, I personally prefer to use *tmap*. 


##Projections

What is the projection of this object?

```{r}
st_crs(Accidents)
```


```{r}
#st_set_crs
#AllCRS <- rgdal::make_EPSG()
#UTM 21 SUR
#32721
#https://epsg.io/32721
#https://www.spatialreference.org/ref/epsg/wgs-84-utm-zone-21s/

Accidents = st_set_crs(Accidents, 32721)
st_crs(Accidents)
```

##Tidyverse in action

```{r}
Accidents %>% names()
```

```{r}
Accidents %>% group_by(ANIO) %>% summarise(Total = n())
```

```{r}
AccidentsData <- Accidents
st_geometry(AccidentsData) <- NULL

AccidentsData %>% group_by(ANIO) %>% summarise(Total = n())
```

```{r}
AccidentsData <- Accidents
AccidentsData %>% group_by(ANIO) %>% summarise(Total = n()) %>% st_set_geometry(NULL)
```

```{r}
Resumen <- AccidentsData %>% group_by(ANIO) %>%
  summarise(Total = n()) %>% 
  st_set_geometry(NULL)
```

```{r}
ggplot(data = Resumen) + geom_bar(aes(x = ANIO, y = Total), stat = "identity") +
  xlab("Años") + ylab("Accidentes") + theme_classic()
```

```{r}
Esquinas <- Accidents %>% group_by(ESQUINA) %>% 
  summarise(Total = n()) %>% 
  st_set_geometry(NULL) %>% 
  arrange(desc(Total))
```

```{r}
Esquinas
```

```{r}
##Improve this plot!

Esquinas %>% filter(Total >60) %>%  
  ggplot() + 
    geom_bar(aes(x=ESQUINA, y = Total), stat = "identity")
```

```{r}
Accidentes2007 <- Accidents %>% filter(ANIO == "2007")
Accidentes2010 <- Accidents %>% filter(ANIO == "2010")

Graves <- Accidents %>% filter(TIPO == "GRAVE")

Graves2008 <- Accidents %>% filter(TIPO == "GRAVE" & ANIO == "2008")


```


#Polygons (States)

```{r}
setwd("~/Dropbox/Teaching/SpatialAnalysis-MontevideoWorkshop2019/Notebooks/Montevideo_Data/Vectoriales_2011")
Zones <- st_read("ine_depto.shp")
```

```{r}
plot(st_geometry(Zones))
```


```{r}
Montevideo <- Zones[Zones$NOMBRE == "MONTEVIDEO",]
plot(st_geometry(Montevideo))
```

```{r}
library(tidyverse)
Montevideo <- Zones %>% filter(NOMBRE == "MONTEVIDEO")
plot(st_geometry(Montevideo))
```

```{r}
st_crs(Montevideo)
st_crs(Zones)

st_crs(Montevideo) == st_crs(Accidentes)
```

```{r}
Montevideo <- st_set_crs(Montevideo,32721)
st_crs(Montevideo) == st_crs(Accidentes)
```

```{r}
plot(st_geometry(Montevideo))
plot(st_geometry(Accidentes), add = TRUE)
```

##The Tidyverse again


#Points in Polygons


#A better map



#Rethinking Points




Vamos a hacer algunos mapas bonitos. 
Creat an animation using 

```{r}
setwd("~/Dropbox/Teaching/SpatialAnalysis-MontevideoWorkshop2019/Notebooks/Montevideo_Data/montevideourbano_shp")

MontevideoUrbano <- st_read("MontevideoUrbano.shp")
```


```{r}

```

<br/>

#Data Sources

All data used in this workflow was downloaded from **Open Source** [pltaforms](https://catalogodatos.gub.uy/dataset?res_format=SHP)

- Data in */montevideorural_shp* and */montevideourbano_shp* were retrieved from the Uruguay [Cátalogo dde datos abiertos](https://catalogodatos.gub.uy/dataset/shapes-del-parcelario-rural-y-urbano).
- Data in */Vectoriales_2011* were retrieved from [Instituto nacional de estadística](http://ine.gub.uy/mapas-vectoriales) 
- Data in */Accidentes2006-2010* were taken from [Catálogo de datos abiertos]((https://catalogodatos.gub.uy/dataset/accidentes-de-transito-montevideo)
- Data in */CyclingMontevideoData* were taken from [Catálogo de datos abiertos](https://catalogodatos.gub.uy/dataset/bicicircuitos-bicicletarios-estaciones-y-talleres-de-reparacion)
Another interesting sources are:

- [geoportal](http://geoportal.mtop.gub.uy/geoservicios)
- [La Encuesta continua de hogares](http://www.ine.gub.uy/web/guest/encuesta-continua-de-hogares1)

